Long-term Water Quality Change in Rhode Island Lakes and Ponds

### 2019-03-07 manuscript/manuscript.Rmd
fresh render with stephen's draft intro

Average yearly scaled temperature in lakes and ponds appear to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 4)`) with the large majority of years with average temperature greater than the long term average occuring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and ponds.", display = "cite")`).  Chlorophyll *a* is also showing an increasing trend over time (slope: `r round(ww_chla_rl$slope,3)` , p-value: `r round(ww_chla_rl$p.value, 4)`) and with the exception of an early high chlorophyll *a* year in 1992 and a slightly above average year in 2003, the above average years have all occurred since 2010 (`r fig_num(name = "chla_ri", caption = "Yearly trend over 20+ years of Chlorphyll a (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).

@@

Average yearly trends for nutrients showed weaker or no trends over time.  Total nitrogen did have a significant trend (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 4)`); however, that trend is driven by the lower than average total nitrogen values in 1993 and 1994 (`r fig_num(name = "tn_ri", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).  Since 1995, the yearly trend is not significant (slope: `r round(ww_tn_rl_no9394$slope,3)`, p-value: `r round(ww_tn_rl_no9394$p.value, 4)`).  Total phosphorus does not show a trend over time in the yearly scaled values (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 4)`) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_ri", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).

@@

In general, broad regional trends were less obvious and in the case of chlorophyll *a* and total phosphorus, the trends were not significant.  In particular, chlorophyll *a* showed a slight, yet non-significant negative trend (slope: `r round(lagos_chla_rl$slope,3)`, p-value: `r round(lagos_chla_rl$p.value, 4)`, `r fig_num(name = "chla_lagos", caption = "Yearly trend over 20+ years of Chlorophyll a (average z-score) in the LAGOSNE states.", display = "cite")`).

@@

Total phosphorus shows a slight and non-significant increasing trend (slope: `r round(lagos_tp_rl$slope, 3)`, p-value: `r round(lagos_tp_rl$p.value,4)`, `r fig_num(name = "tp_lagos", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in the LAGOSNE states.",display = "cite")`) and total nitrogen shows a slight negative non-significant trend (slope: `r round(lagos_tn_rl$slope, 3)`, p-value: `r round(lagos_tn_rl$p.value, 4)`, `r fig_num(name = "tn_lagos", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in the LAGOSNE states.", display = "cite")`).

>>>>

Average yearly scaled temperature in lakes and ponds appear to be increasing (slope: `r round(ww_temp_rl$slope,3)` , p-value: `r round(ww_temp_rl$p.value, 5)`) with the large majority of years with average temperature greater than the long term average occuring in the years since 2000 (`r fig_num(name = "temp_ri", caption = "Yearly trend over 20+ years of lake temperature in Rhode Island lakes and ponds.", display = "cite")`).  Chlorophyll *a* is also showing an increasing trend over time (slope: `r round(ww_chla_rl$slope,3)` , p-value: `r round(ww_chla_rl$p.value, 5)`) and with the exception of an early high chlorophyll *a* year in 1992 and a slightly above average year in 2003, the above average years have all occurred since 2010 (`r fig_num(name = "chla_ri", caption = "Yearly trend over 20+ years of Chlorphyll a (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).

@@

Average yearly trends for nutrients showed weaker or no trends over time.  Total nitrogen did have a significant trend (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`); however, that trend is driven by the lower than average total nitrogen values in 1993 and 1994 (`r fig_num(name = "tn_ri", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).  Since 1995, the yearly trend is not significant (slope: `r round(ww_tn_rl_no9394$slope,3)`, p-value: `r round(ww_tn_rl_no9394$p.value, 5)`).  Total phosphorus does not show a trend over time in the yearly scaled values (slope: `r round(ww_tn_rl$slope,3)` , p-value: `r round(ww_tn_rl$p.value, 5)`) and years that are over or under the average are evenly distributed over the years (`r fig_num(name = "tp_ri", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in Rhode Island lakes and ponds.", display = "cite")`).

@@

In general, broad regional trends were less obvious and in the case of chlorophyll *a* and total phosphorus, the trends were not significant.  In particular, chlorophyll *a* showed a slight, yet non-significant negative trend (slope: `r round(lagos_chla_rl$slope,3)`, p-value: `r round(lagos_chla_rl$p.value, 5)`, `r fig_num(name = "chla_lagos", caption = "Yearly trend over 20+ years of Chlorophyll a (average z-score) in the LAGOSNE states.", display = "cite")`).

@@

Total phosphorus shows a slight and non-significant increasing trend (slope: `r round(lagos_tp_rl$slope, 3)`, p-value: `r round(lagos_tp_rl$p.value,4)`, `r fig_num(name = "tp_lagos", caption = "Yearly trend over 20+ years of Total Phosphorus (average z-score) in the LAGOSNE states.",display = "cite")`) and total nitrogen shows a slight negative non-significant trend (slope: `r round(lagos_tn_rl$slope, 3)`, p-value: `r round(lagos_tn_rl$p.value, 5)`, `r fig_num(name = "tn_lagos", caption = "Yearly trend over 20+ years of Total Nitrogen (average z-score) in the LAGOSNE states.", display = "cite")`).
